Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-07-18 supplement.tex
discussion of OOA parameter fix

\subsection{Estimating two-locus statistics with small sample sizes}

@@

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified parameters that reached
consistent values across all models. These included the timings of recent or
non-central divergences and admixture events:
\begin{itemize}
    \item Neanderthal/human split: 550ka
    \item East/West African population split: 60ka
    \item East African/European split: 50ka
    \item Neanderthal introgression to Eurasian branch: 45ka
    \item Eurasian back-to-African migration: 12ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed (see Table~\ref{tab:fixed-params} for a complete list of all fixed
parameters). This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

@@

@@

\subsubsection{Varying fixed divergence times}

In order to reduce parameter space to avoid overfitting, we fixed parameters
that have previously been inferred, including the timings of the early
back-to-Africa migration contributing to the ancestry of East African
agricultural populations (12ka), the divergence of West and East Africans
(60ka), and the out-of-Africa migration event (50ka). Due to variability in
inference and dating methods, there is some uncertainty surrounding these
dates. We explored how variable timings of these events effect our inferred
models by 1) increasing the fixed times of the E-W African split (to 100ka) and
the out-of-Africa event (to 60ka), and 2) allowing those two parameters to be
jointly fit in model optimization.

>>>>

\subsection{Estimating two-locus statistics with finite sample sizes}

@@

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified some parameters that reached
consistent values across all models. :
\begin{itemize}
    \item Neanderthal/human split: 550ka
    \item Neanderthal introgression to Eurasian branch: 45ka
    \item Eurasian back-to-African migration: 12ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed (see Table~\ref{tab:fixed-params} for a complete list of all fixed
parameters). This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

@@

In our preliminary models, the unconstrained inferred split times between East African populations 
and Europeans was inferred to have occurred around 40kya \sgcomment{check}, which is 
more recent than we may have expected given 
the well-documented presence of modern humans out of Africa at that time\sgcomment{cite}, as 
well as previous genetics-based dating of the split \sgcomment{cite}. 
The inferred split between East and West African populations occurred about $10ka$ earlier. 
We see two non-exclusive explanations for this apparent discrepancy. 
First, there is uncertainty in our model fit of this parameter.
The $95\%$ confidence interval allows for a split as far back as \sgcomment{$XX$}. 
Second, our confidence intervals do not capture all sources of uncertainty. 
For example, model mis-specification can lead to a mismatch between an inferred split time and the first time populations actually split. 
Our model supposes a single OOA event with constant symmetric migrations 
between populations, which is certainly an oversimplification \cite{}. 
Rather than adding the many parameters required to account for these effects, 
we fixed the split time between Europeans and the East African ancestral population at $50ka,$ and
the split between East/West African population at $60ka$. 
We discuss the sensitivity of our model to these assumptions in Section \ref{sec:change_dates}.

@@

\subsubsection{Varying fixed divergence times}
\label{sec:change_dates}
In order to reduce parameter space to avoid overfitting, we fixed parameters
that have previously been inferred, including the timings of the early
back-to-Africa migration contributing to the ancestry of East African
agricultural populations (12ka), the divergence of West and East Africans
(60ka), and the out-of-Africa migration event (50ka). Due to variability in
inference and dating methods, there is some uncertainty surrounding these
dates. We explored how variable timings of these events effect our inferred
models by 1) increasing the fixed times of the E-W African split (to 100ka) and
the out-of-Africa event (to 60ka), and 2) allowing those two parameters to be
jointly fit in model optimization.
